---
title: "WTA Set Analysis"
author: "Peter Tea"
date: "09/07/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Source
Data obtained via Jeff Sackmann's [tennis_wta](https://github.com/JeffSackmann/tennis_wta) repo.

```{r collect_data, echo=FALSE, eval=FALSE}
# WTA set analysis
# How much does winning the 1st set matter in the WTA grand slams?
library(ggplot2)
library(dplyr)

setwd("/Users/petertea/Documents/Github/match-analysis/")
source("/Users/petertea/Documents/Github/match-analysis/src/wta_collect_set_data.R")

# Get data
wta_historic_set_data <- wta_collect_entire_set_data(1968:2019)

# Add win given dropping first set indicator columm:
wta_historic_set_data <- wta_historic_set_data %>%
  mutate(drop_1st_set_and_still_won = ifelse(first_set_winner == 2, 1,0))

# --> Save file..
saveRDS(wta_historic_set_data, file = "wta_historic_set_data.rds")
```



## Any gaps in the data?

Seems like before 1988, we have pretty small data samples.

```{r data_quality, echo = FALSE, message = FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
setwd("/Users/petertea/Documents/GitHub/match-analysis/wta_analysis")
wta_historic_set_data <- readRDS('wta_historic_set_data.rds')

# How many datapoints do we have each year?
wta_historic_set_data %>%
  group_by(year) %>%
  summarise(observations = n()) %>%
  ggplot(aes(x = year, y = observations)) +
  geom_point()
```


## How likely are WTA players to win if they drop the opening set?

Interesting increase from 2001 - 2007, but most players don't end up winning after dropping the opening set.

High proportion in QF and up group could be due to relatively smaller sample sizes.

```{r set_plot, echo = FALSE, message = FALSE, warning=FALSE}
wta_data <- wta_historic_set_data %>%
  filter(year >= 1988) %>%
  #filter(grepl(round, pattern = 'F'))%>% 
  group_by(year) %>%
  summarise(prop = sum(drop_1st_set_and_still_won) / n()) 

wta_data_F <- wta_historic_set_data %>%
  filter(year >= 1988) %>%
  filter(grepl(round, pattern = 'F'))%>% 
  group_by(year) %>%
  summarise(F_prop = sum(drop_1st_set_and_still_won) / n()) 

wta_data_NF <- wta_historic_set_data %>%
  filter(year >= 1988) %>%
  filter(!grepl(round, pattern = 'F'))%>% 
  group_by(year) %>%
  summarise(NF_prop = sum(drop_1st_set_and_still_won) / n()) 

wta_to_plot <- wta_data %>% left_join(wta_data_F, by = 'year') %>%
  left_join(wta_data_NF, by = 'year')

wta_to_plot <- reshape2::melt(wta_to_plot, id=c("year"))

ggplot(data = wta_to_plot, aes(x = year, y = value, color = variable)) +
  geom_point() + geom_line(linetype = 3) + geom_smooth(method='auto', alpha = 0.25) +
  xlab("Year") +
  theme_bw() + 
  ylab("Win Proportion") + 
  ggtitle('Does dropping the 1st Set matter in the WTA')+ 
  scale_color_manual(values=c("blue", "forestgreen", "red"),
                     labels = c('Entire Tournament','QF and up','Before QF'))
```